---
title: "Brauer et al. (2008): Figure 2 Reproduction"
author: "Bin via Codex"
date: "2026-06-18 updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Setup

This analysis reproduces the hierarchically clustered expression matrix in
Figure 2 using the authors' processed data. Code is folded by default and can
be expanded using the **Code** controls.

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

pkgs_cran <- c("pheatmap", "here")
pkgs_bio  <- c("GEOquery")

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

for (p in pkgs_bio)
  if (!requireNamespace(p, quietly = TRUE)) BiocManager::install(p)

for (p in pkgs_cran)
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)

if (!requireNamespace("plotly", quietly = TRUE)) {
  stop("The plotly package is required for the interactive Figure 2 heatmap.")
}

suppressPackageStartupMessages({
  library(GEOquery)
  library(pheatmap)
  library(plotly)
  library(here)
})

i_am("analysis/20260618-Brauer-2008-reanalysis.Rmd")

input_dir  <- here("input")
output_dir <- here("output")
#dir.create(input_dir, recursive = TRUE, showWarnings = FALSE)
#dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

download_if_missing <- function(url, path) {
  if (!file.exists(path)) {
    message("Downloading ", basename(path), " ...")
    utils::download.file(url, path, mode = "wb", quiet = FALSE)
  } else {
    message("Using cached ", basename(path))
  }
}

```

# Acquire processed data and current annotations

The analysis caches the authors' raw TXT and CDT files, the latest available
GPL884 record, and current SGD feature tables under `input/`. Existing files
are never overwritten.

```{r acquire-data}
raw_path <- file.path(input_dir, "dilution_rate_00_raw.txt")
cdt_path <- file.path(input_dir, "dilution_rate_00_raw.cdt")
gpl_path <- file.path(input_dir, "GPL884.soft.gz")
sgd_path <- file.path(input_dir, "SGD_features.tab")
deleted_path <- file.path(input_dir, "deleted_merged_features.tab")

download_if_missing(
  "http://growthrate.princeton.edu/data/dilution_rate_00_raw.txt",
  raw_path
)
download_if_missing(
  "http://growthrate.princeton.edu/data/dilution_rate_00_raw.cdt",
  cdt_path
)
download_if_missing(
  "https://downloads.yeastgenome.org/curation/chromosomal_feature/SGD_features.tab",
  sgd_path
)
download_if_missing(
  "https://downloads.yeastgenome.org/curation/chromosomal_feature/deleted_merged_features.tab",
  deleted_path
)

if (!file.exists(gpl_path)) {
  message("Downloading current GPL884 annotation through GEOquery ...")
  getGEO("GPL884", destdir = input_dir)
}

```

# Reconcile the raw TXT and clustering CDT

The TXT is the authoritative numeric source. The CDT is used to validate row
identity and to recover its historical gene and probe identifiers.

```{r reconcile-inputs}
raw_data <- read.delim(
  raw_path, check.names = FALSE, na.strings = "", quote = "\"",
  stringsAsFactors = FALSE
)
cdt_data <- read.delim(
  cdt_path, check.names = FALSE, na.strings = "", quote = "\"",
  stringsAsFactors = FALSE
)

expr_cols <- grep("^[GNPSLU]0\\.[0-9]+$", names(raw_data), value = TRUE)
stopifnot(
  nrow(raw_data) == 5537L,
  nrow(cdt_data) == 5537L,
  length(expr_cols) == 36L,
  identical(raw_data$GID, cdt_data$GID),
  identical(raw_data$YORF, cdt_data$UID)
)

expr_raw <- as.matrix(raw_data[, expr_cols, drop = FALSE])
expr_cdt <- as.matrix(cdt_data[, expr_cols, drop = FALSE])
storage.mode(expr_raw) <- "numeric"
storage.mode(expr_cdt) <- "numeric"
stopifnot(sum(is.na(expr_raw)) == 866L)

# The CDT contains the same ordered non-missing values, but omits interior
# delimiters in 359 rows. This shifts values left, so it is not a numeric source.
same_value_sequence <- vapply(seq_len(nrow(expr_raw)), function(i) {
  identical(
    as.vector(expr_raw[i, !is.na(expr_raw[i, ])]),
    as.vector(expr_cdt[i, !is.na(expr_cdt[i, ])])
  )
}, logical(1))
stopifnot(all(same_value_sequence))

cdt_position_disagreement <- rowSums(
  xor(is.na(expr_raw), is.na(expr_cdt)) |
    ((!is.na(expr_raw) & !is.na(expr_cdt)) & expr_raw != expr_cdt)
) > 0
stopifnot(sum(cdt_position_disagreement) == 359L)

```

Both files contain **`r nrow(raw_data)` rows** in identical order and the same
ordered non-missing measurements. The raw TXT contains **`r sum(is.na(expr_raw))`
missing values**. The CDT shifts missing-value positions in
**`r sum(cdt_position_disagreement)` rows**, so it is not used as an expression
matrix.

# Map probe IDs through GPL884 and current SGD

Each stable Agilent probe UID is linked to GPL884 and then reconciled against
current SGD features, including deleted and merged-feature mappings. No rows
are discarded based on current annotation status.

```{r map-annotations}
gpl <- getGEO(filename = gpl_path)
gpl_table <- Table(gpl)
gpl_idx <- match(raw_data$YORF, gpl_table$SPOT_ID)
stopifnot(!anyNA(gpl_idx))

sgd_colnames <- c(
  "sgd_id", "feature_type", "qualifier", "feature_name",
  "standard_name", "aliases", "parent_feature", "secondary_sgd_id",
  "chromosome", "start", "stop", "strand", "genetic_position",
  "coordinate_version", "sequence_version", "description"
)
sgd <- read.delim(
  sgd_path, header = FALSE, col.names = sgd_colnames, quote = "",
  comment.char = "", fill = TRUE, stringsAsFactors = FALSE
)
deleted_colnames <- c(
  "old_feature_name", "old_feature_type_status", "chromosome", "start",
  "stop", "strand", "old_sgd_id", "old_secondary_sgd_id",
  "replacement_feature_name", "replacement_sgd_id", "description",
  "notes", "change_date"
)
deleted <- read.delim(
  deleted_path, header = FALSE, col.names = deleted_colnames, quote = "",
  comment.char = "", fill = TRUE, stringsAsFactors = FALSE
)

historical_id <- cdt_data$YORF
direct_idx <- match(historical_id, sgd$feature_name)
deleted_idx <- match(historical_id, deleted$old_feature_name)
replacement_id <- deleted$replacement_feature_name[deleted_idx]
replacement_id[is.na(replacement_id) | replacement_id == ""] <- NA_character_
current_id <- historical_id
current_id[is.na(direct_idx)] <- replacement_id[is.na(direct_idx)]
current_idx <- match(current_id, sgd$feature_name)

mapping_status <- ifelse(
  !is.na(direct_idx), "current",
  ifelse(!is.na(replacement_id), "merged_or_renamed", "unresolved")
)

gene_annotation <- data.frame(
  row_key = raw_data$YORF,
  gid = raw_data$GID,
  probe_uid = raw_data$YORF,
  cdt_historical_id = historical_id,
  gpl_orf = gpl_table$ORF[gpl_idx],
  gpl_feature_name = gpl_table$NAME[gpl_idx],
  gpl_gene_symbol = gpl_table$GENE_SYMBOL[gpl_idx],
  current_sgd_id = sgd$sgd_id[current_idx],
  current_systematic_id = current_id,
  current_standard_name = sgd$standard_name[current_idx],
  current_feature_type = sgd$feature_type[current_idx],
  current_qualifier = sgd$qualifier[current_idx],
  mapping_status = mapping_status,
  stringsAsFactors = FALSE
)

gpl_canonical_id <- ifelse(
  nzchar(gene_annotation$gpl_orf), gene_annotation$gpl_orf,
  gene_annotation$gpl_feature_name
)
gpl_cdt_disagreement <- gpl_canonical_id != historical_id

write.csv(
  gene_annotation,
  file.path(output_dir, "figure2_gene_annotation.csv"),
  row.names = FALSE, na = ""
)

# Stable, unique probe UIDs are retained as row keys even when historical
# features now resolve to the same SGD feature.
rownames(expr_raw) <- gene_annotation$row_key

```

All **`r sum(!is.na(gpl_idx))` / `r nrow(raw_data)`** probes map to GPL884.
There are **`r sum(gpl_cdt_disagreement)`** GPL/CDT identifier disagreements,
and **`r sum(!is.na(current_idx))` / `r nrow(raw_data)`** rows resolve to a
current SGD feature.

```{r annotation-summary, echo=FALSE}
mapping_summary <- as.data.frame(
  table(gene_annotation$mapping_status, useNA = "ifany"),
  stringsAsFactors = FALSE
)
names(mapping_summary) <- c("Mapping status", "Rows")
knitr::kable(mapping_summary, caption = "Current SGD mapping outcomes")

qualifier_summary <- as.data.frame(
  table(gene_annotation$current_qualifier, useNA = "ifany"),
  stringsAsFactors = FALSE
)
names(qualifier_summary) <- c("SGD qualifier", "Rows")
knitr::kable(qualifier_summary, caption = "Current SGD feature qualifiers")
```

The rows with empty "SGD qualifier" are either "blocked reading frame", "transposable element gene", or "pseudogene". These are not excluded from the analysis, as they were present in the original data and may still be biologically relevant.

# Parse and validate sample metadata

Sample metadata are parsed directly from the 36 expression-column names and
validated as the complete six-nutrient by six-dilution-rate design.

```{r sample-metadata}
metadata_match <- regexec("^([GNPSLU])(0\\.[0-9]+)$", expr_cols)
metadata_parts <- regmatches(expr_cols, metadata_match)
stopifnot(all(lengths(metadata_parts) == 3L))

pd_ordered <- data.frame(
  sample = expr_cols,
  nutrient = vapply(metadata_parts, `[`, character(1), 2L),
  dil_rate = as.numeric(vapply(metadata_parts, `[`, character(1), 3L)),
  stringsAsFactors = FALSE,
  row.names = expr_cols
)
pd_ordered$nutrient <- factor(
  pd_ordered$nutrient, levels = c("G", "N", "P", "S", "L", "U")
)

expected_design <- expand.grid(
  nutrient = c("G", "N", "P", "S", "L", "U"),
  dil_rate = c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30),
  stringsAsFactors = FALSE
)
observed_design <- data.frame(
  nutrient = as.character(pd_ordered$nutrient),
  dil_rate = pd_ordered$dil_rate
)
stopifnot(
  !anyNA(pd_ordered),
  !anyDuplicated(observed_design),
  nrow(merge(expected_design, observed_design)) == 36L
)

```

```
knitr::kable(pd_ordered, caption = "Experimental conditions")
```

# Figure 2 — Hierarchical clustering heatmap

All 5,537 rows are clustered using pairwise-complete Pearson correlation and
average-linkage UPGMA. Missing measurements are shown in gray.

```{r cluster-genes}
row_sd <- apply(expr_raw, 1, sd, na.rm = TRUE)
stopifnot(all(is.finite(row_sd)), all(row_sd > 0))

gene_cor <- cor(t(expr_raw), use = "pairwise.complete.obs", method = "pearson")
stopifnot(all(is.finite(gene_cor)))
gene_dist <- as.dist(1 - gene_cor)
gene_clust <- hclust(gene_dist, method = "average")

# Column annotation: nutrient & dilution rate
col_ann <- data.frame(
  Nutrient = pd_ordered$nutrient,
  Rate = factor(pd_ordered$dil_rate),
  row.names = colnames(expr_raw)
)

nutrient_colors <- c(G = "#E41A1C", N = "#377EB8", P = "#4DAF4A",
                     S = "#984EA3", L = "#FF7F00", U = "#A65628")
rate_palette    <- colorRampPalette(c("white","#2C7BB6"))(6)
rate_breaks     <- sort(unique(pd_ordered$dil_rate))

ann_colors <- list(
  Nutrient = nutrient_colors,
  Rate     = setNames(rate_palette, as.character(rate_breaks))
)

# The original figure uses seven integer-centered color bins. Assigning values
# through matrix indexing preserves dimensions (unlike pmax() with a scalar as
# its first argument, which can drop the matrix attributes).
expr_binned <- round(expr_raw)
expr_binned[expr_binned < -3] <- -3
expr_binned[expr_binned > 3] <- 3

# Bright TreeView-style color scale used in the paper:
# green = down, black = 0, red = up. The paper uses log2 ratios
# (sample / glucose-limited 0.25 reference).
heatmap_colors <- c(
  "#00FF00", "#00AA00", "#005500",
  "#000000",
  "#550000", "#AA0000", "#FF0000"
)
heatmap_breaks <- seq(-3.5, 3.5, by = 1)
heatmap_labels <- -3:3

stopifnot(
  identical(dim(expr_binned), dim(expr_raw)),
  identical(dimnames(expr_binned), dimnames(expr_raw)),
  length(heatmap_breaks) == length(heatmap_colors) + 1L,
  length(heatmap_labels) == length(heatmap_colors),
  all(expr_binned[!is.na(expr_binned)] %in% heatmap_labels)
)
```


```{r render-figure-2, include=FALSE}
figure2_path <- file.path(output_dir, "figure2_hierarchical_clustering.png")

pheatmap(
  mat               = expr_binned,
  color             = heatmap_colors,
  breaks            = heatmap_breaks,
  legend_breaks     = heatmap_labels,
  legend_labels     = as.character(heatmap_labels),
  cluster_cols      = FALSE,   # columns pre-ordered by nutrient/rate
  cluster_rows      = gene_clust,
  show_rownames     = FALSE,
  show_colnames     = TRUE,
  fontsize_col      = 7,
  annotation_col    = col_ann,
  annotation_colors = ann_colors,
  border_color      = NA,
  na_col            = "#BDBDBD",
  filename          = figure2_path,
  width             = 8,
  height            = 12,
  main = paste0(
    "Figure 2 — Hierarchical clustering of expression\n",
    "across dilution rates and limiting nutrients\n",
    "(5,537 rows; UPGMA, pairwise-complete Pearson correlation)"
  )
)
```

The interactive heatmap follows the same clustered row order as the saved PNG.
Hover over a cell to identify its gene and view the unbinned expression value.
The axes are fixed so the full heatmap remains in view.

```{r interactive-figure-2, echo=FALSE, fig.width=10, fig.height=11.5, out.width="100%", out.height="1100px"}
plot_row_order <- gene_clust$order
plot_expr_binned <- expr_binned[plot_row_order, , drop = FALSE]
plot_expr_raw <- expr_raw[plot_row_order, , drop = FALSE]
plot_annotation <- gene_annotation[plot_row_order, , drop = FALSE]

plot_standard_name <- plot_annotation$current_standard_name
missing_standard_name <- is.na(plot_standard_name) | plot_standard_name == ""
plot_standard_name[missing_standard_name] <-
  plot_annotation$gpl_gene_symbol[missing_standard_name]
plot_standard_name[
  is.na(plot_standard_name) | plot_standard_name == ""
] <- "(not assigned)"

plot_systematic_id <- plot_annotation$current_systematic_id
missing_systematic_id <- is.na(plot_systematic_id) | plot_systematic_id == ""
plot_systematic_id[missing_systematic_id] <-
  plot_annotation$gpl_orf[missing_systematic_id]
plot_systematic_id[
  is.na(plot_systematic_id) | plot_systematic_id == ""
] <- "(unresolved)"

hover_text <- matrix(
  NA_character_, nrow = nrow(plot_expr_raw), ncol = ncol(plot_expr_raw)
)
for (j in seq_len(ncol(plot_expr_raw))) {
  expression_text <- ifelse(
    is.na(plot_expr_raw[, j]), "NA", sprintf("%.2f", plot_expr_raw[, j])
  )
  hover_text[, j] <- paste0(
    "Standard gene name: ", plot_standard_name,
    "<br>Systematic ID: ", plot_systematic_id,
    "<br>Nutrient: ", as.character(pd_ordered$nutrient[j]),
    "<br>Dilution rate: ", sprintf("%.2f", pd_ordered$dil_rate[j]),
    "<br>Expression value: ", expression_text
  )
}

nutrient_names <- c(
  G = "Glucose", N = "Nitrogen", P = "Phosphate",
  S = "Sulfate", L = "Leucine", U = "Uracil"
)
nutrient_levels <- names(nutrient_names)
nutrient_banner_shapes <- lapply(seq_along(nutrient_levels), function(i) {
  nutrient <- nutrient_levels[i]
  list(
    type = "rect", xref = "paper", yref = "paper",
    x0 = (i - 1) / length(nutrient_levels),
    x1 = i / length(nutrient_levels),
    y0 = 1.01, y1 = 1.055,
    line = list(width = 0),
    fillcolor = unname(nutrient_colors[nutrient])
  )
})
nutrient_banner_labels <- lapply(seq_along(nutrient_levels), function(i) {
  nutrient <- nutrient_levels[i]
  list(
    xref = "paper", yref = "paper",
    x = (i - 0.5) / length(nutrient_levels), y = 1.032,
    text = paste0("<b>", nutrient_names[nutrient], "</b>"),
    showarrow = FALSE,
    font = list(color = "white", size = 12),
    xanchor = "center", yanchor = "middle"
  )
})

bottom_sample_labels <- paste0(
  as.character(pd_ordered$nutrient), " / ",
  sprintf("%.2f", pd_ordered$dil_rate)
)

figure2_interactive <- plot_ly(
  x = colnames(plot_expr_binned),
  y = seq_len(nrow(plot_expr_binned)),
  z = plot_expr_binned,
  type = "heatmap",
  colors = heatmap_colors,
  zmin = -3,
  zmax = 3,
  text = hover_text,
  hoverinfo = "text",
  colorbar = list(
    title = "log2 ratio",
    tickmode = "array",
    tickvals = heatmap_labels,
    ticktext = as.character(heatmap_labels)
  ),
  showscale = TRUE
) |>
  layout(
    title = list(
      text = paste0(
        "Figure 2 — Hierarchical clustering of expression",
        "<br><sup>5,537 rows; UPGMA, pairwise-complete Pearson correlation</sup>"
      )
    ),
    xaxis = list(
      title = "Nutrient / dilution rate",
      side = "bottom",
      tickmode = "array",
      tickvals = colnames(plot_expr_binned),
      ticktext = bottom_sample_labels,
      tickangle = -90,
      fixedrange = TRUE
    ),
    yaxis = list(
      title = "Clustered genes",
      autorange = "reversed",
      showticklabels = FALSE,
      fixedrange = TRUE
    ),
    shapes = nutrient_banner_shapes,
    annotations = nutrient_banner_labels,
    plot_bgcolor = "#BDBDBD",
    height = 1100,
    margin = list(l = 70, r = 90, b = 120, t = 155)
  ) |>
  config(
    scrollZoom = FALSE,
    displayModeBar = FALSE,
    doubleClick = FALSE,
    displaylogo = FALSE
  )

figure2_interactive
```

The reproduced heatmap is saved to
[`output/figure2_hierarchical_clustering.png`](../output/figure2_hierarchical_clustering.png),
and the annotation audit is saved to
[`output/figure2_gene_annotation.csv`](../output/figure2_gene_annotation.csv).

# Identify the phosphate-specific cluster

The paper describes a 133-gene cluster induced specifically during phosphate
limitation. The fixed rule below requires a strong average phosphate signal, a
low non-phosphate background, a minimum phosphate contrast, and consistent
phosphate elevation at every matched dilution rate.

```{r phosphate-specificity}
is_p <- grepl("^P", colnames(expr_raw))
sample_rates <- sub("^[GNPSLU]", "", colnames(expr_raw))

mean_p <- rowMeans(expr_raw[, is_p, drop = FALSE], na.rm = TRUE)
mean_non_p <- rowMeans(expr_raw[, !is_p, drop = FALSE], na.rm = TRUE)
p_effect <- mean_p - mean_non_p

p_cols <- which(is_p)
p_consistency_count <- vapply(seq_len(nrow(expr_raw)), function(i) {
  sum(vapply(p_cols, function(j) {
    same_rate_non_p <- which(!is_p & sample_rates == sample_rates[j])
    is.finite(expr_raw[i, j]) &&
      expr_raw[i, j] > mean(expr_raw[i, same_rate_non_p], na.rm = TRUE)
  }, logical(1)))
}, integer(1))

# Fixed, interpretable thresholds on the log2 expression-ratio scale.
phosphate_fixed <- (
  mean_p >= 1 &
  mean_non_p <= 0.5 &
  p_effect >= 0.75 &
  p_consistency_count == length(p_cols)
)

stopifnot(sum(phosphate_fixed) == 133L)

# Average each gene across the six dilution rates for each nutrient.
nutrient_codes <- c("G", "N", "P", "S", "L", "U")
nutrient_means <- vapply(nutrient_codes, function(nutrient) {
  rowMeans(
    expr_raw[, grepl(paste0("^", nutrient), colnames(expr_raw)), drop = FALSE],
    na.rm = TRUE
  )
}, numeric(nrow(expr_raw)))
colnames(nutrient_means) <- nutrient_codes

fixed_rows <- which(phosphate_fixed)
# Ordering is for presentation only; membership is determined entirely by the
# fixed thresholds above.
fixed_rows <- fixed_rows[order(p_effect[fixed_rows], decreasing = TRUE)]

fixed_standard_name <- gene_annotation$current_standard_name[fixed_rows]
missing_standard <- is.na(fixed_standard_name) | fixed_standard_name == ""
fixed_standard_name[missing_standard] <-
  gene_annotation$gpl_gene_symbol[fixed_rows][missing_standard]

fixed_systematic_id <- gene_annotation$current_systematic_id[fixed_rows]
missing_systematic <- is.na(fixed_systematic_id) | fixed_systematic_id == ""
fixed_systematic_id[missing_systematic] <-
  gene_annotation$gpl_orf[fixed_rows][missing_systematic]

phosphate_gene_table <- data.frame(
  `Probe UID` = gene_annotation$probe_uid[fixed_rows],
  `Standard gene name` = fixed_standard_name,
  `Systematic ID` = fixed_systematic_id,
  `Feature type` = gene_annotation$current_feature_type[fixed_rows],
  Qualifier = gene_annotation$current_qualifier[fixed_rows],
  Glucose = round(nutrient_means[fixed_rows, "G"], 3),
  Nitrogen = round(nutrient_means[fixed_rows, "N"], 3),
  Phosphate = round(nutrient_means[fixed_rows, "P"], 3),
  Sulfate = round(nutrient_means[fixed_rows, "S"], 3),
  Leucine = round(nutrient_means[fixed_rows, "L"], 3),
  Uracil = round(nutrient_means[fixed_rows, "U"], 3),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

phosphate_output_path <- file.path(
  output_dir, "figure2_phosphate_specific_candidates.csv"
)
write.csv(
  phosphate_gene_table,
  phosphate_output_path,
  row.names = FALSE,
  na = ""
)
```

The fixed rule identifies **`r sum(phosphate_fixed)` genes**, reproducing the
reported cluster size. Matching the number 133 alone does not establish that
its membership is identical to the manually identified dendrogram branch in
the paper. The table is ordered by decreasing mean phosphate-versus-other
nutrient contrast; this ordering does not affect membership.

```{r phosphate-gene-table, echo=FALSE}
knitr::kable(
  phosphate_gene_table,
  caption = paste0(
    "The 133 fixed-threshold phosphate-specific genes. Expression columns ",
    "are means across the six dilution rates for each nutrient."
  ),
  align = c("l", "l", "l", "l", "l", rep("r", 6))
)
```

The fixed-rule candidates show the expected phosphate-specific block across
the six matched dilution rates.

```{r phosphate-cluster-heatmap, fig.width=8, fig.height=8, out.width="85%"}
phosphate_expr <- expr_raw[phosphate_fixed, , drop = FALSE]
phosphate_cor <- cor(
  t(phosphate_expr), use = "pairwise.complete.obs", method = "pearson"
)
stopifnot(all(is.finite(phosphate_cor)))
phosphate_clust <- hclust(as.dist(1 - phosphate_cor), method = "average")

pheatmap(
  mat = phosphate_expr,
  color = heatmap_colors,
  breaks = heatmap_breaks,
  legend_breaks = heatmap_labels,
  legend_labels = as.character(heatmap_labels),
  cluster_cols = FALSE,
  cluster_rows = phosphate_clust,
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize_col = 7,
  annotation_col = col_ann,
  annotation_colors = ann_colors,
  border_color = NA,
  na_col = "#BDBDBD",
  main = "133 phosphate-specific candidates (fixed rule)"
)
```

The same 133-gene annotation and nutrient-mean table is available as
[`output/figure2_phosphate_specific_candidates.csv`](../output/figure2_phosphate_specific_candidates.csv).

## Why PHO84 is absent from the authors' processed matrix

`PHO84` (`YML123C`; probe `A_06_P5417`) is present on GPL884 and in the
original GSE8825 arrays, but it is absent from `dilution_rate_00_raw.txt` and
therefore from the 5,537 rows clustered above. To investigate this discrepancy,
the official GEO raw archive was downloaded from the
[GSE8825 supplementary-file endpoint](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE8825&format=file).
The archive and its 36 gzip-compressed Agilent Feature Extraction files are
cached under [`input/`](../input/); their provenance is recorded in
[`input/README.md`](../input/README.md).

The following code reads the PHO84 record directly from each original GSM raw
file. The sorted GSM accessions correspond to the 36 expression columns in the
GEO series matrix and in the authors' matrix.

```{r investigate-pho84}
geo_raw_dir <- file.path(input_dir, "GSE8825_RAW")
geo_raw_files <- sort(list.files(
  geo_raw_dir, pattern = "^GSM[0-9]+\\.txt\\.gz$", full.names = TRUE
))
stopifnot(length(geo_raw_files) == length(expr_cols))

read_agilent_probe <- function(path, probe_id) {
  raw_lines <- readLines(gzfile(path), warn = FALSE)
  header_line <- grep("^FEATURES\tFeatureNum\t", raw_lines)[1]
  probe_line <- grep(
    paste0("\t", probe_id, "\t"), raw_lines,
    fixed = TRUE
  )
  stopifnot(length(header_line) == 1L, length(probe_line) == 1L)

  header <- strsplit(raw_lines[header_line], "\t", fixed = TRUE)[[1]]
  values <- strsplit(raw_lines[probe_line], "\t", fixed = TRUE)[[1]]
  as.data.frame(
    as.list(stats::setNames(values, header)),
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}

pho84_raw <- do.call(rbind, lapply(
  geo_raw_files, read_agilent_probe, probe_id = "A_06_P5417"
))
pho84_raw$GSM <- sub("\\.txt\\.gz$", "", basename(geo_raw_files))
pho84_raw$condition <- expr_cols

numeric_raw_fields <- c(
  "LogRatio", "gProcessedSignal", "rProcessedSignal",
  "gIsWellAboveBG", "rIsWellAboveBG", "gIsSaturated", "rIsSaturated",
  "gIsFeatNonUnifOL", "rIsFeatNonUnifOL",
  "gIsFeatPopnOL", "rIsFeatPopnOL", "IsManualFlag"
)
pho84_raw[numeric_raw_fields] <- lapply(
  pho84_raw[numeric_raw_fields], as.numeric
)

pho84_above_bg <- pho84_raw$condition[pho84_raw$gIsWellAboveBG == 1]
stopifnot(
  identical(pho84_above_bg, expr_cols[grepl("^P", expr_cols)]),
  !"A_06_P5417" %in% raw_data$YORF,
  all(pho84_raw$LogRatio[grepl("^P", pho84_raw$condition)] == 2)
)

pho84_quality_summary <- data.frame(
  Check = c(
    "Green channel above background",
    "Red channel above background",
    "Saturated in either channel",
    "Non-uniform feature outlier in either channel",
    "Population outlier in either channel",
    "Manual flag"
  ),
  Arrays = c(
    sum(pho84_raw$gIsWellAboveBG),
    sum(pho84_raw$rIsWellAboveBG),
    sum(pho84_raw$gIsSaturated | pho84_raw$rIsSaturated),
    sum(pho84_raw$gIsFeatNonUnifOL | pho84_raw$rIsFeatNonUnifOL),
    sum(pho84_raw$gIsFeatPopnOL | pho84_raw$rIsFeatPopnOL),
    sum(pho84_raw$IsManualFlag)
  )
)
knitr::kable(
  pho84_quality_summary,
  caption = "PHO84 quality flags in the 36 original Agilent arrays"
)
```

The green channel passes Agilent's `gIsWellAboveBG` test in exactly
**`r sum(pho84_raw$gIsWellAboveBG)` of 36 arrays**: the six phosphate-limited
arrays (**`r paste(pho84_above_bg, collapse = ", ")`**). The red channel is
above background in **`r sum(pho84_raw$rIsWellAboveBG)` arrays**. There is no
saturation, non-uniform-feature, population-outlier, or manual flag for PHO84.

GEO's sample metadata identifies channel 1/green/Cy3 as the shared
glucose-limited reference culture at dilution rate 0.25, and
channel 2/red/Cy5 as the nutrient-limited experimental culture. Thus red—not
green—is the starvation/limitation sample. In all six phosphate arrays,
PHO84 reaches the Feature Extraction ceiling of `LogRatio = 2`, corresponding
to a 100-fold red-to-green ratio, or `log2(100) = 6.644` in the GEO series
matrix.

The authors' retained matrix has at most
**`r max(rowSums(is.na(expr_raw)))` missing values per row**. PHO84's weak green
reference signal in 30 of 36 arrays is therefore strongly consistent with its
removal by an intensity/background-completeness filter before construction of
the 5,537-row matrix. This is an inference from the original measurements and
quality flags; without the authors' preprocessing code, the precise exclusion
rule cannot be established definitively.

# Figure 3 — deferred

Figure 3 is intentionally held for later work. Its previous R implementation
is retained below inside an HTML comment and is not evaluated by knitr.

<!--
if (FALSE) { # Deferred: Figure 3 is intentionally disabled.
cat("\nComputing SVD for Figure 3...\n")

# Center each gene (row) before SVD, as is standard
expr_centered <- t(scale(t(expr_imp), center = TRUE, scale = FALSE))

# SVD: X = U D V^T
# Eigengenes are the rows of V^T (columns of V), scaled by singular values
svd_res <- svd(expr_centered)

# Fraction of variance explained by each eigengene
eigenvalues   <- svd_res$d^2
frac_variance <- eigenvalues / sum(eigenvalues)

cat("Fraction variance — top 8 eigengenes:\n")
print(round(frac_variance[1:8], 4))
# Paper reports eigengene 1 = 43.6% of variation

# Build data frame of eigengene projections
# Columns of V (right singular vectors) = eigengene patterns across samples
n_eig   <- 8   # show top 8 in panel B; use 4 for panel C traces
V_mat   <- svd_res$v  # nrow = 36 samples, ncol = 36

eigengene_df <- as.data.frame(V_mat[, 1:n_eig])
colnames(eigengene_df) <- paste0("EG", 1:n_eig)
eigengene_df$sample    <- colnames(expr_imp)
eigengene_df$nutrient  <- pd_ordered$nutrient
eigengene_df$dil_rate  <- pd_ordered$dil_rate
eigengene_df$nutrient_fac <- factor(pd_ordered$nutrient,
                                    levels = c("G","N","P","S","L","U"))

# ── Panel A: heatmap of first 36 eigengenes (the V matrix) ────
# Shows all 36 eigengenes × 36 samples (Figure 3A in paper)

png("figure3_svd.png", width = 3200, height = 2600, res = 200)
layout(matrix(c(1,1,2,3,3,3,3), nrow = 1), widths = c(2,0.6,3.4))

# Panel A — eigengene heatmap (samples × eigengenes)
par(mar = c(5,4,4,1))
image(
  1:36, 1:36,
  t(V_mat[, 36:1]),   # flip so EG1 at top
  col    = colorRampPalette(c("#008000","black","#CC0000"))(100),
  xlab   = "Sample index",
  ylab   = "Eigengene",
  main   = "A  36 eigengenes",
  axes   = FALSE
)
axis(1, at = seq(1,36,6), labels = seq(1,36,6), cex.axis=0.8)
axis(2, at = c(1,18,36), labels = c("36","18","1"), cex.axis=0.8, las=1)

# Panel B — eigenvalue bar chart
par(mar = c(5,4,4,1))
barplot(
  frac_variance[1:8],
  names.arg = paste0("EG",1:8),
  horiz     = TRUE,
  col       = c("#CC0000", rep("gray60", 7)),
  border    = NA,
  xlab      = "Fraction of variance",
  main      = "B  Eigenvalues",
  cex.names = 0.75,
  cex.axis  = 0.75,
  xlim      = c(0, 0.5)
)
text(frac_variance[1], 7.5,
     paste0(round(frac_variance[1]*100,1),"%"),
     pos=4, cex=0.75, col="darkred")

# Panel C — expression of 4 top eigengenes across nutrient / rate
# Replicate the 6-panel grid (one column per nutrient) × 4 rows (eigengenes)
par(mar = c(0,0,0,0))  # reset before grid below
dev.off()   # close first; redo with separate png for cleaner layout

# ── Panel C as its own clean multi-panel plot ──────────────────
# For each eigengene (1-4) × each nutrient (G,N,P,S,L,U):
#   plot eigengene projection vs. dilution rate

nutrients_ordered <- c("G","N","P","S","L","U")
nut_labels        <- c(G="Glucose", N="Nitrogen", P="Phosphate",
                       S="Sulfate", L="Leucine", U="Uracil")

rates_all <- sort(unique(pd_ordered$dil_rate))  # 0.05…0.30

make_eigengene_panel <- function(eg_idx, nutrient_code,
                                 ylim = c(-0.35, 0.10)) {
  sub_df <- eigengene_df %>%
    filter(nutrient == nutrient_code) %>%
    arrange(dil_rate)
  
  y_col <- paste0("EG", eg_idx)
  y_vals <- sub_df[[y_col]]
  
  ggplot(sub_df, aes(x = dil_rate, y = .data[[y_col]])) +
    geom_line(color = "black", linewidth = 0.6) +
    geom_point(shape = 1, size = 2, color = "black") +
    scale_x_continuous(breaks = rates_all,
                       labels = ifelse(eg_idx == 4,
                                       as.character(rates_all), "")) +
    scale_y_continuous(limits = ylim,
                       breaks = pretty(ylim, n=3)) +
    labs(
      x = if (eg_idx == 4) nut_labels[nutrient_code] else "",
      y = if (nutrient_code == "G") paste0("Eigengene ", eg_idx) else ""
    ) +
    theme_classic(base_size = 8) +
    theme(
      axis.text.x  = if (eg_idx == 4) element_text(angle=45, hjust=1,
                                                   size=6) else element_blank(),
      axis.ticks.x = if (eg_idx == 4) element_line() else element_blank(),
      axis.text.y  = element_text(size = 6),
      plot.margin  = unit(c(0.5,0.5,0.5,0.5), "mm")
    )
}

# Adaptive y-limits per eigengene
ylims <- list(
  EG1 = range(eigengene_df$EG1) * 1.1,
  EG2 = range(eigengene_df$EG2) * 1.1,
  EG3 = range(eigengene_df$EG3) * 1.1,
  EG4 = range(eigengene_df$EG4) * 1.1
)

panels <- list()
for (eg in 1:4) {
  for (nut in nutrients_ordered) {
    panels[[paste0("EG",eg,"_",nut)]] <-
      make_eigengene_panel(eg, nut, ylim = ylims[[paste0("EG",eg)]])
  }
}

panel_grid <- gridExtra::arrangeGrob(
  grobs  = panels,
  ncol   = 6,
  nrow   = 4,
  top    = "C  Expression levels of 4 most significant eigengenes"
)

# ── Panel B — eigenvalue barplot as ggplot ────────────────────
frac_df <- data.frame(
  eigengene = factor(paste0("EG",1:8), levels = paste0("EG",8:1)),
  variance  = frac_variance[1:8]
)

panel_B <- ggplot(frac_df, aes(x = variance, y = eigengene)) +
  geom_col(fill = c(rep("grey60",7),"#CC0000"), width = 0.7) +
  geom_text(aes(label = ifelse(eigengene == "EG1",
                               paste0(round(variance*100,1),"%"), "")),
            hjust = -0.1, size = 2.5, color = "darkred") +
  scale_x_continuous(expand = expansion(mult=c(0,0.15)),
                     labels = scales::percent_format(accuracy=1)) +
  labs(x = "Fraction of variance", y = NULL,
       title = "B  Eigengene weights (eigenvalues)") +
  theme_classic(base_size = 8) +
  theme(axis.text.y = element_text(size = 7))

# ── Assemble & save Figure 3 ──────────────────────────────────
cat("Saving Figure 3...\n")

# Full figure 3 — panels B and C side by side
fig3 <- gridExtra::arrangeGrob(
  panel_B,
  panel_grid,
  ncol   = 2,
  widths = c(1, 3.5),
  top    = "Figure 3 — SVD decomposition of expression data\n(Brauer et al. 2008 reproduction)"
)

ggsave("figure3_svd_panels_BC.png", fig3,
       width = 14, height = 8, dpi = 200, bg = "white")
cat("  → figure3_svd_panels_BC.png\n")

# ── 6. Quick summary of eigengene 1 vs growth rate ────────────
cat("\n── Correlation of Eigengene 1 with dilution rate ──\n")
cat("  R² =", round(cor(eigengene_df$EG1, eigengene_df$dil_rate)^2, 3),
    " (paper reports R² = 0.69)\n")

cat("\nDone! Output files:\n")
cat("  figure2_hierarchical_clustering.png\n")
cat("  figure3_svd_panels_BC.png\n")
} # End deferred Figure 3 block.
-->
